RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-11-06 doc/paper.tex
Conclusions and Abstract revised

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
          This paper solves the well-known prominent problem of Symbolic regression. This problem have a wide range of applications.  For example, the obtained SR models could serve as insight towards  establishing  physical theory behind, thus helping establishing reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology,  modeling analytic representations of the exciton binding energy in the energy science, and many applications in the material science. We propose a new meta-heuristic based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are a several concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to determine suitable coefficients in the models (solutions);  ($iii$) a novel fitness function is utilized which combines accuracy score, error score, and the size of model into one, more balanced value. The experiments are conducted on two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} to 14 other competitors from the literature under different levels of noise in the input data.  Our method obtains new best exact percentages on the both benchmark sets comparing to the literature approaches. Moreover, new unbiased randomly generated set of instances are introduced and scalability and robustness of our method is demonstrated on these instances in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

  This paper deals with solving the well--known Symbolic regression (SR)  problem. This problem is a generalization of more interdisciplinary appearing  linear and logistic regression problems. SR problem have been intensively studied in the literature over the last 15 years, where a few dozens of algorithms were constructed, mainly based on Genetic programming and on the application of different kinds of neural networks. We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. Our approach is based on the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the oridinary least square method to efficiently determine internal coefficients of the candidate solutions,  partially. Another key aspect is the use of a carefully constructed fitness function, which appears as  a combination of three important scores: RMSE, $R^2$  scores and the size of model (solution).  Last but not least, the neighbor of solution in the local search is carefully chosen that suits to small modifications of a three representing solutions (models). \textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from literature on the two ground truth benchmark sets from literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed efficiently of our method which was able to produce the best average ranking results in terms of exact percentage wh comparison to all other competitors on the two ground-truth benchmark sets, in presence of the three different levels of noise in the input data. In all cases, new best exact percentages are obtained by means of \textsc{Rils-Rols} method. In some cases, these differences are statistically significant in favor of our \textsc{Rils-Rols} method. Moreover, scalability and robustness of this method is  checked and confirmed on unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of model, the number of variables, average runtimes, etc.  \fxnote{TODO: case-study about large-scale data to be added}

  In the future work, one could think of constructing a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics. Replacing ILS scheme of \textsc{Rils-Rils} with, a general scheme Variable neighborhood search meta-heuristic is a reasonable option to try out. Additionally, as a proof of the efficiency of our algorithm, our tool should be provided to various research communities out of the are of Computer science, such as the ones from physics or chemistry to help for  seeking new hypotheses or to obtain valuable insights on their experimental evaluations.

>>>>

\author[1]{Aleksandar Kartelj}
\author[2]{Marko Djukanovi\'c}
	\address[1]{$kartelj@matf.bg.ac.rs$, \\  Faculty of Mathematics, University of Belgrade, Serbia}
    \address[2]{$ marko.djukanovic@pmf.unibl.org$,\\   Faculty of Natural Sciences and Mathematics, University of Banja Luka, Bosnia and Herzegovina}
	\begin{abstract}
          In this paper we solve the well-known, prominent problem of Symbolic regression, intensively studied over the last two decades. This problem have a wide range of applications;  for example, the obtained SR models could serve theoretically, providing insight towards  establishing  physical theory behind, thus helping establishing reasonable hypotheses.  Practical applications include revealing complex ecological dynamics in ecology, discovering mutations effects on protein stability in biology,  modeling analytic representations of the exciton binding energy in the energy science, finding models for band gap of NaCl-type compounds
         in material science, just to name a few. We propose a new meta-heuristic based approach, called \textsc{Rils}-\textsc{Rols} algorithm to solve the problem of Symbolic regression. There are a several key concepts utilized within this algorithm: ($i$) Iterated local search scheme is used as the backbone; ($ii$) ordinary least square method is used within the local search procedure to make it efficient and to fit the coefficients in models (solutions);  ($iii$) a novel fitness function is utilized which combines an accuracy score, error score, and the size of model into one, more compact score. The experiments are conducted on the two ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz} comparing \textsc{Rils}-\textsc{Rols} to 14 other competitors from the literature considering different levels of noise in the input data.  Our method obtains new best exact percentages on the both benchmark sets comparing to the literature approaches. Moreover, new unbiased randomly generated set of problem instances are introduced where scalability and robustness of our method is thoroughly demonstrated in terms of various instance properties. 
	\end{abstract}
	\maketitle

@@

  In this paper we dealt with solving the well--known Symbolic regression (SR)  problem. SR problem is a generalization of more interdisciplinary appearing  linear and logistic regression problems. %SR problem have been intensively studied in the literature over the last 15 years, where a few dozens of algorithms were constructed, mainly based on Genetic programming and on the application of different kinds of neural networks. 
   We proposed a meta-heuristic approach, called \textsc{Rils}-\textsc{Rols}. Our approach is built upon   the Iterated local search (ILS) scheme. The first crucial concept within this scheme is utilizing the ordinary least square method to efficiently determine internal coefficients that fit the candidate solution,  thus in this way  the search process is focused on considering more promising solutions. Another key aspect is the utilization of a carefully constructed fitness function, which appears as  a combination of three important scores: RMSE, $R^2$  scores and the size of model (solution).  Last but not least, a carefully constructed local search is applied systematically exploring solutions obtained as 1--perturbations of the considered solution.  1--perturbations of a solution are based on applying the six per-node change operations on the solution's tree representation.

    \textsc{Rils-Rols} method is compared to the 14 other competitor algorithms from literature on the two ground truth benchmark sets from literature, namely \textsc{Feynman} and \textsc{Strogatz}.  Additionally, we introduced a non-biased benchmark set of randomly generated instances, labelled by \textsc{Random}.  The experimental evaluation confirmed efficiently of our method which produced the best average ranking results in terms of exact percentage when compared to all other competitors on the two ground-truth benchmark sets, in presence of all three different levels of noise in the input data. In all cases, new best exact percentages are obtained by means of \textsc{Rils-Rols} method. In some cases, these differences are  statistically significant in favor of our \textsc{Rils-Rols} method. Moreover, scalability and robustness of this method is  checked and confirmed on the unbiased \textsc{Random} benchmark set in terms of a few different criteria such as the size of model, the number of variables in exact models, and average runtimes.

  In the future work, one could think of constructing a hybrid of \textsc{Rils-Rols} method with some other meta-heuristics to further boost quality of the  obtained results. For example, replacing ILS scheme of \textsc{Rils-Rils} with  a more general Variable neighborhood search (VNS) scheme is a reasonable option to try out. VNS   utilizes more stronger and systematic system of solutions diversification  and thus   escaping from local optima could be much easier.  Additionally, as a proof of efficiency and benefits of our algorithm, this tool could  be applied to solving problems from various problem domains out of the areas of computer science, for example from physics or chemistry, to help for setting up hypotheses or to obtain valuable insights on   experimental evaluations.
